---
title: "EDA of two datasets for cox regression"
author: "Hermela Shimelis"
format: html
editor: visual
date: last-modified
date-format: long
---

## Exploratory data analysis

## I. **Framingham heart study dataset**

The Framingham Heart Study is a longituidinal prospective study of the etiology of cardiovascular disease among subjects in the community of Framingham, Massachusetts.

```{r echo=F, message=FALSE}
# load library
library(readr)
library(dplyr)
```

```{r}

# Import data
df <- read_csv("~/git_directory/IDC6940_hazardous-conditions/Datasets/framingham.csv")

```

```{r}
# view data
head(df,10)


# view data structure 
str(df) # all variables are numerical

# check missing data
sum(is.na(df)) # 645 Nas data

```

Evaluate continuous variables

```{r}
hist(df$age)
hist(df$cigsPerDay)
hist(df$totChol)
hist(df$sysBP)
hist(df$diaBP)
hist(df$BMI)
hist(df$heartRate)
hist(df$glucose)

```

## II. NCCTG lung cancer data: R dataset from "Survival" package

The dataset Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

#### Column descriptions

|            |                                                                                                                                                                                                    |
|:---------------|:-------------------------------------------------------|
| inst:      | Institution code                                                                                                                                                                                   |
| time:      | Survival time in days                                                                                                                                                                              |
| status:    | censoring status 1=censored, 2=dead                                                                                                                                                                |
| age:       | Age in years                                                                                                                                                                                       |
| sex:       | Male=1 Female=2                                                                                                                                                                                    |
| ph.ecog:   | ECOG performance score as rated by the physician. 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed \<50% of the day, 3= in bed \> 50% of the day but not bedbound, 4 = bedbound |
| ph.karno:  | Karnofsky performance score (bad=0-good=100) rated by physician                                                                                                                                    |
| pat.karno: | Karnofsky performance score as rated by patient                                                                                                                                                    |
| meal.cal:  | Calories consumed at meals                                                                                                                                                                         |
| wt.loss:   | Weight loss in last six months (pounds)                                                                                                                                                            |
|            |                                                                                                                                                                                                    |

```{r}
# Vetrans' admistration lung cancer study
library(survival)

data(cancer, package="survival")

head(cancer)
```

```{r}

# view data structure 
str(cancer) # all variables are numerical

# check missing data
sum(is.na(cancer)) # 67 Nas data
```

```{r}
library(dplyr)
unique(cancer$inst)
n_distinct(cancer$inst) # the data was collected from 18 institutions, and an unknown institution.

```

\- Examine hwo many patients were sampled from each institution

```{r}

# Number of patients from each institution:
ct <- cancer%>%group_by(as.factor(inst))%>%summarise(count = n())
print(ct,20)

# exclude unknown institution
cancer <- cancer%>%filter(!is.na(inst))
```

Insight: At least two patients were included from each institution. One institution with a single patient is excluded from the data since the institution is unknown.

Evaluate continious variables

```{r}
hist(cancer$time)
hist(cancer$age)
hist(cancer$ph.karno)
hist(cancer$pat.karno)
hist(cancer$meal.cal)
hist(cancer$wt.loss)


```

Insight: most of the variables are fairly normally distributed. Meal calorie appear to be right skewed.

```{r}
status <- cancer%>%group_by(status)%>%summarise(count = n())
print(status)

sex <- cancer%>%group_by(sex)%>%summarise(count = n())
print(sex)

phecog <- cancer%>%group_by(ph.ecog)%>%summarise(count = n())
print(phecog)

```

Insight: ph.ecog variable has only 4 category, one of which is represented by only one individual. Remove two rows, one with NA category and another with single observation.

```{r}
# Remove a row with category 3 and NA
cancer <- cancer%>%filter(!is.na(ph.ecog))
cancer <- cancer%>%filter(ph.ecog != 3)
```
